---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# finalfourr

<!-- badges: start -->
<!-- badges: end -->

The goal of finalfourr is to provide an accessible and powerful framework for 
analyzing NCAA men's basketball, with a primary focus on predicting 
Sweet Sixteen teams and supporting deeper insights through simulation, 
comparison, and visualization tools.

```{r setup, include=FALSE}
library(finalfourr)
```

## Installation Instructions:

In order to install the finalfourr package you must first install the cbbdata
package then install finalfourr:

```{r, eval=FALSE}
devtools::install_github("andreweatherman/cbbdata")
devtools::install_github("ZayneHMaughan/finalfourr")
```

Secondly, you need an API key to query data from the cbbdata package. 
```{r, eval=FALSE}

cbbdata::cbd_create_account(username, email, password, confirm_password)

```

## Core Functionality 

`predict_s16_teams(year)`

Predicts which teams would make the Sweet Sixteen using an XGBoost model trained on season statistics from 2008–2023.

* Uses SMOTE to address class imbalance.

* Returns a ranked list of 16 predicted teams.

* Example:

```{r predictions-example}
predict_s16_teams(2020)
```

## Data Exploration Tools

### Search Teams by Keyword
```{r}
search_team("utah", include_years = TRUE)
```

### Compare Two Teams
```{r compare-example}
comparison <- compare_teams(2021, "Gonzaga", "Baylor")
comparison
```
```{r plot-example, fig.width=8, fig.height=5}
plot_compare_teams(comparison)
```

### Get Teams by Conference
```{r}
conf_teams <- get_teams_by_conf(2018)
names(conf_teams) # List of conference names
head(conf_teams$`B10`) # Teams in the Big Ten
```

## Simulations and Utilities

### Simulate Matchup
```{r}
set.seed(385)
simulate_matchup("Utah", "Baylor", 2015)
```

### Filter Team Stats

```{r}
filter_team("N.C. State", season = 2016)
```


## Conclusion

The finalfourr package enables analysts, data scientists, and fans to:

* Predict NCAA tournament outcomes.

* Explore team performance trends.

* Simulate realistic game scenarios.

It offers both rigorous modeling and exploratory tools to deepen your understanding of NCAA men’s basketball.

