---
title: "Finalfourr Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(finalfourr)
```

Package can be found at [https://github.com/ZayneHMaughan/finalfourr](https://github.com/ZayneHMaughan/finalfourr)


# Introduction

Every spring, NCAA Division I men’s college basketball teams compete in a single-elimination tournament known as March Madness. The tournament features 68 teams, selected based on regular season performance and conference championships. These teams are seeded into four regions, and through a series of rounds—First Four, First Round (Round of 64), Second Round (Round of 32)—they compete to advance to later stages:

* Sweet Sixteen (final 16 teams)

* Elite Eight (final 8)

* Final Four (final 4)

* National Championship

The Sweet Sixteen represents a key milestone in the tournament. Teams that make it this far have already won two games and are often considered among the strongest performers of the season.

The goal of the finalfourr package is to predict which teams will reach the Sweet Sixteen based on seasonal team statistics. While the original intent of the model was to predict the Final Four or even the national champion, these outcomes proved to be too variable and difficult to model accurately. Through trial and error, we found that predicting Sweet Sixteen teams is a more feasible and reliable problem. This stage balances the unpredictability of early upsets with the signal of sustained team performance, making it a strong target for statistical modeling.

Although Sweet Sixteen prediction is the central focus, the package also includes functions for simulating matchups, comparing team stats, and exploring historical NCAA data.



## Installation Instructions:

In order to install the finalfourr package you must first install the cbbdata
package then install finalfourr:

```{r, eval=FALSE}
devtools::install_github("andreweatherman/cbbdata")
devtools::install_github("ZayneHMaughan/finalfourr")
```

Secondly, you need an API key to query data from the cbbdata package. 
```{r, eval=FALSE}

cbbdata::cbd_create_account(username, email, password, confirm_password)

```

---

## Predicting Sweet 16

### `predict_s16_teams()`

This returns a table with the teams that are predicted to make it to the sweet 
16. To make the predictions we created a predictive model based on seasonal 
statistics for years 2008 - present. We are training on years 2008 - 2023. We 
used all variables except for the number of games played, the number of wins, 
the number of losses, and the number of game minutes. There were a bunch of
variables that are included in the model so we will attach a the variable 
importance of this model. This model is a XGBoosted tree model that is a binary 
classifier. The response of this model is simply an indicator of whether or not 
a team made it to the Sweet 16. A challenge that comes from this is that because 
there are over 300 teams in division one men's basketball, having only sixteen 
in the desired class creates imbalanced classes. To combat this SMOTE was done 
on the data to try and achieve greater class balance. The goal is to predict on
a year's data so it is a deterministic model. The probabilities given by a 
prediction come from this model's prediction on it's confidence they would make
it to the sweet sixteen.

For this example, we will be using the 2020 season to make a prediction as to
what teams would have made it if their season was not ended before the tournament
due to Covid.

```{r predictions-example}
predict_s16_teams(2020)
```
This output shows the predicted 16 teams that our model determines who would 
have made it to the Sweet 16 for the 2020 season.

Because there were around 80 variables that were trained on, the variables most
important to the prediction are as follows:
```{r variable-importance}
caret::varImp(model)
```

This vignette also demonstrates how to use several functions from the package to 
explore NCAA basketball team data, including:

- Comparing two teams (`compare_teams()` and `plot_compare_teams()`)
- Listing bracket teams for a given year (`get_bracket_teams()`)
- Getting team lists by conference (`get_teams_by_conf()`)
- Searching for teams by keyword (`search_team()`)
- Simulating a matchup between two teams (`simulate_matchup()`)
- Flipping a coin between two teams (`flip_a_coin()`)

All functions assume the presence of a dataset like `cbb_data`, containing 
columns such as `year`, `team`, `CONF`, and `POSTSEASON`.

--- 


## Comparing Two Teams

### `compare_teams()`

Returns offensive and defensive metrics for two teams in a given year:

- `ftr`: Free throw rate  
- `two_pt_pct`: 2-point percentage  
- `three_pt_pct`: 3-point percentage  
- `def_ftr.y`: Opponent free throw rate  
- `def_two_pt_pct`: Opponent 2-point percentage  
- `def_three_pt_pct`: Opponent 3-point percentage

```{r compare-example}
comparison <- compare_teams(2021, "Gonzaga", "Baylor")
comparison
```

### `plot_compare_teams()`

Visualizes the comparison in a bar chart.

```{r plot-example, fig.width=8, fig.height=5}
plot_compare_teams(comparison)
```

### `plot_compare_teams_line()`

Visualizes the comparison in two line charts.

```{r plot-example2, fig.width=8, fig.height=5}
plot_compare_teams_line(comparison)
```

---

## Getting Bracket Teams

### `get_bracket_teams()`

Returns all teams that made the NCAA tournament in a given year. Teams are 
identified via the `POSTSEASON` column.

```{r get-bracket-example}
bracket_teams <- get_bracket_teams(2023)
head(bracket_teams)
```

**Note**: Any `POSTSEASON` value not equal to `"0"`, `"N/A"`, or `NA` is 
considered bracket-eligible.

---

## Getting Teams by Conference

### `get_teams_by_conf()`

Returns a named list of conferences and the teams that belonged to each in a
given year.

```{r get-conf-example}
conf_teams <- get_teams_by_conf(2018)
names(conf_teams) # List of conference names
head(conf_teams$`B10`) # Teams in the Big Ten
```

---

### Searching for Teams by Keyword using `search_team()`

`search_team()` allows you to quickly find teams whose names contain a specific
keyword. This is useful when you're working with this large dataset and want to 
filter down to teams of interest and available years of data.

For example, we can see all teams with the keyword `utah`:

```{r}
search_team("utah", include_years = TRUE)
```

---

### Simulating a Matchup Between Two Teams using `simulate_matchup()`

`simulate_matchup()` allows you to simulate head-to-head matchups 
between two NCAA basketball teams for a given season. This simulation is based 
on KenPom-style metrics, including each team's adjusted offensive and 
defensive efficiency, tempo, and additional stats that reflect shooting and 
ball control (like effective field goal percentage and turnover rate).

Unlike deterministic models, this function uses a probabilistic Monte Carlo 
approach, running a large number of simulated games (e.g., 1,000) to estimate 
win probabilities. In each simulation, team scores are generated using a 
normal distribution, where the mean is based on expected points 
(from offensive and defensive ratings), and the standard deviation reflects 
team consistency (based on shooting and turnover metrics). After generating 
scores for both teams in a simulated game, the model simply compares the two 
scores and awards the win to the team with the higher score. This process is 
repeated across many simulations, and the team that wins the majority of 
games is considered the predicted winner.

This approach introduces randomness to mimic the natural variability 
in real games, allowing for more realistic and nuanced predictions. This tool 
is particularly useful for comparing teams across conferences, evaluating 
tournament matchups, or simply exploring “what if” scenarios with 
statistical rigor.

For example, here we simulate 1000 games between Utah and Baylor in the 2015 
season:

```{r}
set.seed(387)
simulate_matchup("Utah", "Baylor", 2015)
```

---

### Flipping a Coin Between Two Teams using `flip_a_coin()`

`flip_a_coin()` is a whimsical tool that stochastically selects a winner 
between two teams—essentially a virtual coin flip. It provides no analytical 
value but can be a fun way to break ties or help decide between picking two 
teams.

For example, here we flip a coin between BYU and Utah:

```{r, results='hold'}
set.seed(348)
flip_a_coin("BYU", "Utah")
```

---

## Notes

- These functions rely on consistent naming in the dataset: columns like `year`,
`team`, `CONF`, and `POSTSEASON` must be present.
- Use validation helpers 
(e.g., `validate_inputs()`, `validate_bracket_inputs()`) to ensure robust 
function calls.

### Get All of the Team Stats Using `filter_team()`

`filter_team()` allows you to pull of the available stats for a team during a 
a given season. 

```{r}
filter_team("N.C. State", season = 2016)
```
This returns a tibble of all available information for North Carolina State
University in 2016.

## Notes
- If season is Null filter team will return all of the teams stats for all 
available seasons. 

---

### Plot the Power Ranking Using `power_plot()`

`plot_power()` allows you to visualize how a specific college basketball team's 
power ranking—measured by the barthag metric—has evolved over time. This plot is especially useful for identifying trends, such as periods of sustained success 
or decline, and  comparing team trajectories across seasons. The only input to 
the function is  the team name (as a string), and it returns a base R plot 
showing the team's power rating from 2013 to the most recent season available 
in the dataset.

```{r ,fig.width=8, fig.height=5}
plot_power("Utah St.")
```
This plot displays Utah State having a very low power ranking from the year 
2013 and decreasing to their worst power ranking in 2018. The team then sky 
rockets to a power ranking of about 0.83 in 2019 and have maintained a high 
ranking since that time. 

## Conclusion

The finalfourr package offers a comprehensive toolkit for exploring and 
modeling NCAA men's basketball performance, with a central focus on predicting 
which teams are likely to reach the Sweet Sixteen. Our 
predict_s16_teams() function leverages an XGBoost binary 
classifier, trained on historical season statistics and enhanced with class 
balancing techniques like SMOTE, to deliver data-driven predictions 
with strong performance.

Beyond prediction, finalfourr provides several utility functions for 
simulating matchups, comparing teams, and retrieving tournament and 
conference-specific data. Tools like simulate_matchup() utilize Monte 
Carlo methods to account for natural game randomness, while functions 
like compare_teams() and plot_power() support exploratory data analysis 
and visual storytelling.

Overall, this package is a valuable resource for analysts, 
sports fans, and data scientists interested in NCAA basketball modeling, 
enabling both rigorous analysis and casual exploration.
