---
title: "finalfourr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{finalfourr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message= FALSE, warning=FALSE}
library(finalfourr)
```

## Introduction

This vignette demonstrates how to use several functions from the package to explore NCAA basketball team data, including:

- Comparing two teams (`compare_teams()` and `plot_compare_teams()`)
- Listing bracket teams for a given year (`get_bracket_teams()`)
- Getting team lists by conference (`get_teams_by_conf()`)
- Searching for teams by keyword (`search_team()`)
- Simulating a matchup between two teams (`simulate_matchup()`)
- Flipping a coin between two teams (`flip_a_coin()`)

All functions assume the presence of a dataset like `cbb_data`, containing columns such as `year`, `team`, `CONF`, and `POSTSEASON`.

---

## Comparing Two Teams

### `compare_teams()`

Returns offensive and defensive metrics for two teams in a given year:

- `ftr`: Free throw rate  
- `two_pt_pct`: 2-point percentage  
- `three_pt_pct`: 3-point percentage  
- `def_ftr.y`: Opponent free throw rate  
- `def_two_pt_pct`: Opponent 2-point percentage  
- `def_three_pt_pct`: Opponent 3-point percentage

```{r compare-example}
comparison <- compare_teams(2021, "Gonzaga", "Baylor")
comparison
```

### `plot_compare_teams()`

Visualizes the comparison in a bar chart.

```{r plot-example, fig.width=8, fig.height=5}
plot_compare_teams(comparison)
```

---

## Getting Bracket Teams

### `get_bracket_teams()`

Returns all teams that made the NCAA tournament in a given year. Teams are identified via the `POSTSEASON` column.

```{r get-bracket-example}
bracket_teams <- get_bracket_teams(2023)
head(bracket_teams)
```

**Note**: Any `POSTSEASON` value not equal to `"0"`, `"N/A"`, or `NA` is considered bracket-eligible.

---

## Getting Teams by Conference

### `get_teams_by_conf()`

Returns a named list of conferences and the teams that belonged to each in a given year.

```{r get-conf-example}
conf_teams <- get_teams_by_conf(2018)
names(conf_teams)      # List of conference names
head(conf_teams$`Big Ten`)  # Teams in the Big Ten
```

---

### Searching for Teams by Keyword using `search_team()`

`search_team()` allows you to quickly find teams whose names contain a specific keyword. This is useful when you're working with this large dataset and want to filter down to teams of interest and available years of data.

For example, we can see all teams with the keyword `utah`:

```{r}
search_team("utah", include_years = TRUE)
```

---

### Simulating a Matchup Between Two Teams using `simulate_matchup()`

`simulate_matchup()` allows you to simulate head-to-head matchups between two teams in a given season. This can help estimate win probabilities based on adjusted offensive, defensive, and tempo metrics.

For example, here we simulate 1000 games between Utah and Baylor in the 2015 season:

```{r}
set.seed(387)
simulate_matchup("Utah", "Baylor", 2015)
```

---

### Flipping a Coin Between Two Teams using `flip_a_coin()`

`flip_a_coin()` is a whimsical tool that stochastically selects a winner between two teams—essentially a virtual coin flip. It provides no analytical value but can be a fun way to break ties or help decide between picking two teams.

For example, here we flip a coin between BYU and Utah:

```{r, results='hold'}
set.seed(348)
flip_a_coin("BYU", "Utah")
```

---

## Notes

- These functions rely on consistent naming in the dataset: columns like `year`, `team`, `CONF`, and `POSTSEASON` must be present.
- Use validation helpers (e.g., `validate_inputs()`, `validate_bracket_inputs()`) to ensure robust function calls.

## Session Info

```{r session-info}
sessionInfo()
```

